---
title: "Calibration Plot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration Plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

```{r setup}

library(predtools)
library(magrittr)
library(dplyr)
library(ggplot2)

```

## What is calibration plot?

Calibration plot is a visual tool to assess the agreement between predictions and observations in different percentiles (mostly deciles) of the predicted values.


`calibration_plot` function constructs calibration plots based on provided predictions and observations columns of a given dataset. Among other options implemented in the function, one can evaluate prediction calibration according to a grouping factor (or even from multiple prediction models) in one calibration plot.


## A step-by-step guide.
Imagine the variable y indicates risk of disease recurrence in a unit of time. We have a prediction model that quantifies this risk given a patient's age, disease severity level, sex, and whether the patient has a comorbidity. 

The package comes with two exemplary datasets. `dev_data` and `val_data`. We use the dev_data as the development sample and the `val_data` as the external validation sample.

```{R}

data(dev_data)
data(val_data)

```


`dev_data` has `r dim(predtools::dev_data)[1]` rows. `val_data` has `r dim(predtools::val_data)[1]` rows. 

Here are the first few rows of `dev_data`:

```{R echo=FALSE}

knitr::kable(dev_data[1:7,])

```


We use the development data to fit a logistic regression model as our risk prediction model:


```{R}

reg <- glm(y~sex+age+severity+comorbidity,data=dev_data,family=binomial(link="logit"))
summary(reg)

```

Given this, our risk prediction model can be written as:

```{R echo=FALSE}

cfs <- coefficients(reg)
str<-paste0(round(cfs[1],4),"+",paste0(round(cfs[-1],4),"*",names(cfs[-1]),collapse="+"))
str_risk_model <- gsub("+-", "-", str, fixed = T)

```

$\bf{ logit(p)=`r str_risk_model`}$.


Now, we can create the calibration plot in development and validation datasets by using `calibration_plot` function.
```{R}

dev_data$pred <- predict.glm(reg, type = 'response')
val_data$pred <- predict.glm(reg, newdata = val_data, type = 'response')

calibration_plot(data = dev_data, obs = "y", pred = "pred", title = "Calibration plot for development data")
calibration_plot(data = val_data, obs = "y", pred = "pred", y_lim = c(0, 0.6),
                 title = "Calibration plot for validation data", group = "sex")

```
