---
title: "Intercept Adjustment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intercept Adjustment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

```{r setup}

library(predtools)
library(magrittr)
library(dplyr)
library(ggplot2)

```

## What is intercept adjustment?

In clinical prediction modeling, model updating refers to the practice of modifying a prediction model before it is used in a new setting to improve its performance. One of the simplest updating methods for risk predication models is a fixed odds-ratio transformation of predicted risks to improve the model’s calibration-in-the-large.


`interceptAdj` function uses an approximate equation for recovering the conditional odds-ratio from the observed mean and predicted variance of risks in validation and development sets, respectively.


## A step-by-step guide.
Imagine the variable y indicates risk of disease recurrence in a unit of time. We have a prediction model that quantifies this risk given a patient's age, disease severity level, sex, and whether the patient has a comorbidity. 

The package comes with two exemplary datasets. `dev_data` and `val_data`. We use the dev_data as the development sample and the `val_data` as the external validation sample.

Model updating matters when there is a considerable difference between mean of the observed risks in development and validation sets. The average of y in the above two datasets are almost identical. Therefore, to have a meaningful scenario, we create a secondary (arbitrary) outcome (y_alt) in `val_data` with a lower average (by ~ 50%).

```{R}

data(dev_data)
data(val_data)

```

```{R echo=FALSE}
set.seed(1)
val_data$y_alt <- ifelse(val_data$y == 0, 0,
                         ifelse(runif(n = nrow(val_data)) <= 0.5, 0, 1))

val_data %>%
  select(y, y_alt) %>%
  summary() %>%
  knitr::kable()

```

`dev_data` has `r dim(predtools::dev_data)[1]` rows. `val_data` has `r dim(predtools::val_data)[1]` rows. 

Here are the first few rows of `dev_data`:

```{R echo=FALSE}

knitr::kable(dev_data[1:7,])

```


We use the development data to fit a logistic regression model as our risk prediction model:


```{R}

reg<-glm(y~sex+age+severity+comorbidity,data=dev_data,family=binomial(link="logit"))
summary(reg)

```

Given this, our risk prediction model can be written as:

```{R echo=FALSE}

cfs <- coefficients(reg)
str<-paste0(round(cfs[1],4),"+",paste0(round(cfs[-1],4),"*",names(cfs[-1]),collapse="+"))
str_risk_model <- gsub("+-", "-", str, fixed = T)

```

$\bf{ logit(p)=`r str_risk_model`}$.


First, let's see the calibration plot in development and validation datasets. We use `calibration_plot` from our package to create calibration plots.
```{R}

dev_data$pred <- predict.glm(reg, type = 'response')
val_data$pred <- predict.glm(reg, newdata = val_data, type = 'response')

calibration_plot(data = dev_data, obs = "y", pred = "pred", title = "Calibration plot for development data")
calibration_plot(data = val_data, obs = "y_alt", pred = "pred", y_lim = c(0, 0.6),
                 title = "Calibration plot for validation data")

```


To adjust the predicted risks for the validation set, we estimate the correction factor by using function `odds_adjust`:
```{R}

odds_correction_factor <- odds_adjust(p0 = mean(dev_data$y), p1 = mean(val_data$y_alt), v = var(dev_data$pred))
odds_correction_factor

```

We can now recalibrate the predictions and reproduce the calibration plot for the validation set.
```{R}

dev_data$pred <- predict.glm(reg, type = 'response')
val_data$pred <- predict.glm(reg, newdata = val_data, type = 'response')

val_data$odds_adj <- (val_data$pred / (1 - val_data$pred)) * odds_correction_factor
val_data$pred_adj <- val_data$odds_adj / (1 + val_data$odds_adj)

val_data$id <- c(1 : nrow(val_data))
val_data_long <- reshape(data = val_data, direction = "long", varying = c("pred", "pred_adj"), v.name = "preds",
                         idvar = "id", timevar = "Method", times = c("Premitive", "Adjusted"))

calibration_plot(data = val_data, obs = "y_alt", pred = "pred_adj",
                 title = "Calibration plot for development data - after recalibration")
calibration_plot(data = val_data_long, obs = "y_alt", pred = "preds", group = "Method",
                 title = "Calibration plot for development data - before and after recalibration")
```

